---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# substrait

<!-- badges: start -->
[![R-CMD-check](https://github.com/paleolimbot/substrait/workflows/R-CMD-check/badge.svg)](https://github.com/paleolimbot/substrait/actions)
[![Codecov test coverage](https://codecov.io/gh/paleolimbot/substrait/branch/master/graph/badge.svg)](https://app.codecov.io/gh/paleolimbot/substrait?branch=master)
<!-- badges: end -->

The goal of substrait is to provide an R interface to the [Substrait](https://substrait.io) cross-language serialization for relational algebra. This is an experimental package that is under heavy development!

## Installation

You can install the development version of substrait from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("paleolimbot/substrait")
```

## Usage

You can create Substrait objects using the `substrait` base object or using `substrait_create()`:

```{r example}
library(substrait)

substrait$Type$Boolean$create()
substrait_create("substrait.Type.Boolean")
```

You can convert an R object *to* a Substrait object using `as_substrait(object, type)`:

```{r}
(msg <- as_substrait(list(i8 = list()), "substrait.Type"))
```

The `type` can be either a string of the qualified name or an object (which is needed to communicate certain types like `"substrait.Expression.Literal.Decimal"` which has a `precision` and `scale` in addition to the `value`).

Restore an R object *from* a Substrait object using `from_substrait(message, prototype)`:

```{r}
from_substrait(msg, list())
```

Substrait objects are list-like (i.e., methods defined for `[[` and `$`), so you can get or set fields. Note that unset is different than `NULL` (just like an R list).

```{r}
msg$i8$nullability <- substrait$Type$Nullability$NULLABILITY_NULLABLE
msg
```

The constructors are currently implemented using about 1000 lines of auto-generated code made by inspecting the nanopb-compiled .proto files and RProtoBuf. This is probably not the best final approach but allows us to get started writing good `as_substrait()` and `from_substrait()` methods for various types of R objects.

## Under the hood

Substrait objects are represented as a classed version of the `raw()` vector containing serialized data. This makes the representation independent from the protocol buffer library use to encode/deocde the objects and is particularly useful with `expect_identical()`.

```{r}
unclass(msg)
```

Currently, the [RProtoBuf](https://cran.r-project.org/package=RProtoBuf) package is used to serialize and deserialize objects. The RProtoBuf package provides excellent coverage of the protocol buffer API and is well-suited to general use; however, the Substrait package only uses a small portion of these features. The package registers the .proto files from Substrait (of which this package has an internal copy) so that you can use RProtobuf to read, modify, and write messages independent of the constructors in the package.

```{r}
(msg_rpb <- RProtoBuf::P("substrait.Type")$read(unclass(msg)))
as_substrait(msg_rpb)
```

An approach more suited to the Substrait use-case is [nanopb](https://github.com/nanopb/nanopb), which is vendorable (i.e., no linking to system libraries) and supports all the protocol buffer features used by Substrait. The compiled .proto encoders/decoders and vendored library is currently included in the `src/` directory but only one encoder is implemented:

```{r}
bytes <- substrait:::r_encode_substrait_Type_Boolean(
  type_variation_reference = 233,
  nullablity = substrait$Type$Nullability$NULLABILITY_REQUIRED
)

cat(RProtoBuf::P("substrait.Type.Boolean")$read(bytes)$toString())
```

There are probably other approaches, but most of them will require some type of auto-generated wrapper code and the portability and compile speed of nanopb is hard to beat.
