---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# substrait

<!-- badges: start -->
[![R-CMD-check](https://github.com/paleolimbot/substrait/workflows/R-CMD-check/badge.svg)](https://github.com/paleolimbot/substrait/actions)
<!-- badges: end -->

The goal of substrait is to provide an R interface to the [Substrait](https://substrait.io) cross-language serialization for relational algebra.

## Installation

You can install the development version of substrait from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("paleolimbot/substrait")
```

You will need the [protocol buffers](https://developers.google.com/protocol-buffers) development files installed. You can do this on MacOS using `brew install protobuf`, on Debian/Ubuntu using `apt-get install libprotobuf-dev`, and on Red Hat/Fedora by installing `protobuf-devel`.

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(substrait)
library(RProtoBuf)

files <- list.files(
  system.file("substrait/proto", package = "substrait"),
  "\\.proto$", recursive = TRUE
)

readProtoFiles2(
  files,
  protoPath = system.file("substrait/proto", package = "substrait")
)

(raw_type_boolean <- substrait:::r_encode_substrait_Type_Boolean(
  type_variation_reference = 10,
  nullablity = 2
))

d <- P("substrait.Type.Boolean")
message <- d$read(raw_type_boolean)
cat(message$toString())
```
