---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# substrait

<!-- badges: start -->
<!-- badges: end -->

The goal of substrait is to provide an R interface to the [Substrait](https://substrait.io) cross-language serialization for relational algebra.

## Installation

You can install the development version of substrait from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("paleolimbot/substrait")
```

You will need the [protocol buffers](https://developers.google.com/protocol-buffers) development files installed. You can do this on MacOS using `brew install protobuf`, on Debian/Ubuntu using `apt-get install libprotobuf-dev`, and on Red Hat/Fedora by installing `protobuf-devel`.

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(substrait)
## basic example code
```
